Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated

### 2016-12-16 ms_pre_post.Rmd
Add some progress to Discussion; more general stuff moved forward

Is evolution of one costly inbreeding strategy inhibited by another non-costly alternative strategy?
----------------------------------------------

When either pre-copulatory or post-copulatory inbreeding strategy allele values were fixed to zero, the alternative inbreeding strategy evolved to inbreeding avoidance even when costly (Figure 3A,C). However, when one inbreeding strategy could evolve and had zero cost, the other strategy initially evolved towards inbreeding avoidance but then back towards random choice across generations following increasing evolution of inbreeding avoidance (i.e., negative allele values) through the former (zero cost) strategy (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ value in the last $10000$ generations was $-0.868$ (Figure 3A), meaning that a female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to mate with a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. However, when $F_{a}$ allele values could also evolve, grand mean $M_{a}$ allele value in the last $10000$ generations was only $-0.288$ (Figure 3B). Further, given fixed $M_{a}=0$, grand mean $F_{a}$ value in the last $10000$ generations was $-1.092$ (Figure 3C), meaning that a female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, when $M_{a}$ values could also evolve and $P_{a}$ values were negative, grand mean $F_{a}$ value in the last $10000$ generations was only $-0.070$ (Figure 3D). Evolution of post-copulatory inbreeding avoidance facilitated evolution of polyandry (Figure 3B,C), while monandry evolved in the absence of post-copulatory inbreeding avoidance (Figure 3A,D; polyandry was cost free in all these simulations). Overall, evolution of a costly inbreeding strategy that readily evolved in isolation was greatly impeded by evolution of a less costly strategy.

Is evolution of one inbreeding strategy precluded by existence of another already adaptive strategy
----------------------------------------------

When polyandry and post-copulatory inbreeding strategy allele values were fixed for adaptive inbreeding avoidance, pre-copulatory inbreeding avoidance evolved (i.e., $M_{a}$ values became increasingly negative; Figure 4A). Such evolution still occurred, but to a much smaller degree, when pre-copulatory inbreeding avoidance was costly (Figure 4B). However, after $40000$ generations, $M_{a}$ values were less negative when post-copulatory inbreeding avoidance and polyandry were fixed at non-zero values than when they also evolved from initial values of zero (e.g., compare the blue lines in Figures 4A versus Figure 1A).

When pre-copulatory inbreeding strategy values were fixed for adaptive inbreeding avoidance, mean $F_{a}$ allele values did not consistently trend toward negative values (Figure 4C,D). Additionally, mean $P_{a}$ allele values consistently decreased over generations regardless of the cost of post-copulatory inbreeding strategy (Figure 4C,D). Consequently, when $c_{F}=0$, $F_{a}$ allele values had no effect because females were almost exclusively monandrous, resulting in high drift of $F_{a}$ values among replicates (illustrated by wide standard errors, Figure 4C). However, when $c_{F}=0.02$, $F_{a}$ values remained near zero to minimise direct costs. This lack of evolution of post-copulatory inbreeding avoidance was driven by a lack of polyandry, and therefore an inability of females to bias fertilisation among multiple mates. When pre-copulatory inbreeding avoidance and polyandry were both fixed ($M_{a}=-10$ and $P_{a}=1$), $F_{a}$ allele values evolved to similarly negative means as $M_{a}$ allele values in Figure 4A,B (see Supporting Information).

@@

We have used *in silico* experiments to test three key hypotheses concerning evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry. The modelling that we develop provides a novel framework for approaching empirical studies of inbreeding avoidance, using simulations to address relevant knowledge gaps in understanding evolution of polyandry and inbreeding avoidance through pre-copulatory and post-copulatory mechanisms. Accordingly, we modelled evolution of three separate traits affecting polyandry and inbreeding avoidance, including tendency for polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy.

@@

The opportunity to adjust inbreeding has been widely considered as a driver of adaptive evolution of polyandry [@Tregenza2002; @Foerster2003; @Akcay2007; @Varian-Ramos2012; @Kingma2013; @Lehtonen2015; @Reid2014]. Recently, @Duthie found that the conditions under which selection for polyandry and pre-copulatory inbreeding avoidance is predicted are highly restrictive. However, @Duthie assumed that fertilisation of offspring was unbiased with respect to mate kinship, meaning that females could not express post-copulatory choice. In contrast, we found that when post-copulatory inbreeding avoidance could evolve, selection for polyandry was greatly strengthened (Figure 1). The proposition that polyandry might facilitate cryptic female choice among males of varying compatibility is not new [e.g., @Zeh1997], but the existence of selection for post-copulatory inbreeding avoidance causing feedback to select for polyandry has not been formally modelled. Our model therefore has widespread implications for future studies of evolution of polyandry.

@@

General predictions
-------------------------------------------------------------------

Some broad predictions are possible given the verbal theory that we have developed and the populations that we have simulated. For example, given that post-copulatory inbreeding avoidance is useless for monandrous females and most effective for polyandrous females, while pre-copulatory inbreeding avoidance is perhaps most effective for monandrous females and less effective for polyandrous females, it would be interesting to test whether or not occurrence of post-copulatory inbreeding avoidance covaries positively, and pre-copulatory inbreeding avoidance covaries negatively, with degree of polyandry. In addition to polyandry making post-copulatory inbreeding avoidance relevant, polyandry might also covary positively with post-copulatory inbreeding avoidance due to the feed back that post-copulatory inbreeding avoidance has on facilitating evolution of polyandry itself. Hence, well-grounded biological hypotheses might be derived for the relationship between female multiple mating and mode of inbreeding avoidance. To test such hypotheses, more work is needed to identify the degree to which females of different species engage in polyandrous and the degree to which females express both pre-copulatory and post-copulatory inbreeding avoidance.

>>>>

How do cost asymmetries affect long-term persistence of reproductive strategies?
----------------------------------------------

When either pre-copulatory ($M_{a}$) or post-copulatory ($F_{a}$) inbreeding avoidance allele values were fixed to zero, the alternative inbreeding strategy evolved to inbreeding avoidance even when costly (Figure 3A,C). Females therefore evolved to avoid inbreeding, and thereby avoid the cost of producing inbred offspring, through whichever route was available.

However, when both pre-copulatory and post-copulatory inbreeding avoidance could evolve, their relative evolutionary dynamics depended on their relative costs. When one inbreeding strategy had zero cost, the other strategy initially evolved towards inbreeding avoidance (i.e., negative $M_{a}$ or $F_{a}$) but then evolved back towards random mating or fertilisation (i.e., $M_{a}=0$ or $F_{a}=0$) across generations following increasing evolution of inbreeding avoidance through the cost-free strategy (Figure 3B,D). <!--- For example, given fixed $F_{a}=0$, grand mean $M_{a}$ value in the last $10000$ generations was $-0.868$ (Figure 3A), meaning that a female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to mate with a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. However, when $F_{a}$ allele values could also evolve, grand mean $M_{a}$ allele value in the last $10000$ generations was only $-0.288$ (Figure 3B). Further, given fixed $M_{a}=0$, grand mean $F_{a}$ value in the last $10000$ generations was $-1.092$ (Figure 3C), meaning that a female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a high cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, when $M_{a}$ values could also evolve and $P_{a}$ values were negative, grand mean $F_{a}$ value in the last $10000$ generations was only $-0.070$ (Figure 3D). --->

Evolution of post-copulatory inbreeding avoidance facilitated evolution of polyandry (Figure 3B,C), while monandry evolved in the absence of post-copulatory inbreeding avoidance (Figure 3A,D; polyandry was cost free in all these simulations). <!--- Overall, evolution of a costly inbreeding strategy that readily evolved in isolation was greatly impeded by evolution of a less costly strategy. --->

How is evolution of one inbreeding strategy affected by the existence of another already adaptive strategy?
----------------------------------------------

When polyandry alleles ($P_{a}$) were fixed to be positive so that all females were expected to mate multiply and post-copulatory inbreeding allele ($F_{a}$) values were fixed for adaptive inbreeding avoidance, pre-copulatory inbreeding avoidance evolved (i.e., $M_{a}$ values became increasingly negative; Figure 4A). Such evolution still occurred, but to a much smaller degree, when pre-copulatory inbreeding avoidance was costly (Figure 4B). However, after $40000$ generations, $M_{a}$ values were less negative when post-copulatory inbreeding avoidance and polyandry were fixed at non-zero values than when they also evolved from initial values of zero (e.g., compare the blue lines in Figures 4A versus Figure 1A).

Conversely, when pre-copulatory inbreeding allele ($M_{a}$) values were fixed for adaptive inbreeding avoidance, mean $F_{a}$ allele values did not consistently become negative over generations (Figure 4C,D). However, in this case, mean $P_{a}$ allele values consistently decreased over generations, reflecting selection against polyandry regardless whether or not post-copulatory inbreeding strategy was costly (Figure 4C,D). Consequently, when $c_{F}=0$, $F_{a}$ allele values had no effect because females were almost exclusively monandrous, resulting in high drift of $F_{a}$ values among replicates (illustrated by wide standard errors, Figure 4C). However, when $c_{F}=0.02$, $F_{a}$ values remained near zero to minimise direct costs. This lack of evolution of post-copulatory inbreeding avoidance was driven by a lack of polyandry, and therefore an inability of females to bias fertilisation among multiple mates. When pre-copulatory inbreeding avoidance and polyandry were both fixed ($M_{a}=-10$ and $P_{a}=1$), $F_{a}$ allele values evolved to similarly negative means as $M_{a}$ allele values in Figure 4A,B (see Supporting Information).

@@

Evolution of reproductive strategies cannot be presumed to occur in isolation in wild populations, meaning that observed patterns of reproductive strategy within and across taxa are likely to be affected by evolutionary feedbacks and (potentially asymmetric) costs. We have addressed key questions regarding the effect of such feedbacks and costs on evolution of pre-copulatory inbreeding avoidance, post-copulatory inbreeding avoidance, and polyandry. A critical analysis of our model results in new predictions concerning evolutionary dynamics given the joint evolution of these traits, which may be used to formulate new empirical hypotheses within and across taxa.

For example, our results highlight a simple but theoretically under-developed conclusion concerning the relationship between evolution of female mating frequency and pre-copulatory versus post-copulatory inbreeding avoidance. Post-copulatory inbreeding avoidance is useless for monandrous females, and most effective for highly polyandrous females that can choose amongst multiple mates to ensure that offspring are fertilised favourably. In contrast, pre-copulatory inbreeding avoidance is perhaps most effective for monandrous females and less effective for polyandrous females. This conclusion is reflected in our simulation results, as selection for post-copulatory inbreeding avoidance was negligible in monandrous populations, resulting in high drift of allele values over generations due to the inability of females to express post-copulatory choice (e.g., Figure 4C,D). The effect of monandry on selection for pre-copulatory inbreeding avoidance was weaker, but pre-copulatory inbreeding avoidance was generally stronger when populations where monandrous (e.g., Figure 3A vs 3B; see also Supporting Information p. S1-1). In addition to polyandry increasing selection for post-copulatory inbreeding avoidance, polyandry might also covary positively with post-copulatory inbreeding avoidance due to the feed back that post-copulatory inbreeding avoidance has on facilitating evolution of polyandry itself, as observed in our model (Figure 1). It would therefore be interesting to test the hypothesis that, across taxa, occurrence of post-copulatory inbreeding avoidance covaries positively, and pre-copulatory inbreeding avoidance covaries negatively, with degree of polyandry. To test this hypothesis, more work is needed to identify the degree to which females of different species engage in polyandrous and the degree to which females express both pre-copulatory and post-copulatory inbreeding avoidance.

Another general inference from our model concerns the dynamics of phenotypes with highly overlapping effects with respect to selection, realised through functionally different traits. Pre-copulatory inbreeding avoidance, typically expressed through mating behaviour, versus post-copulatory inbreeding avoidance, typically expressed through phenotypes affecting fertilisation, are both driven by the shared selective pressure of inbreeding depression in offspring viability. Hence, evolution of each of these strategies might make the other strategy more or less adaptive by modifying its impact on the degree to which a female avoids inbreeding and consequently has offspring that express inbreeding depression.

@@

The opportunity to adjust inbreeding has been widely considered as a driver of adaptive evolution of polyandry [@Tregenza2002; @Foerster2003; @Akcay2007; @Varian-Ramos2012; @Kingma2013; @Lehtonen2015; @Reid2014]. We found that when post-copulatory inbreeding avoidance could evolve, selection for polyandry was greatly strengthened (Figure 1). The proposition that polyandry might facilitate cryptic female choice among males of varying compatibility is not new [e.g., @Zeh1997], but our model provides support for this verbal hypothesis and therefore has widespread implications for future studies of evolution of polyandry.

@@
